---
title: "Homework 7" 
# subtitle: "possible subtitle goes here"
author:
  - Xiaokang Liu
date: "`r format(Sys.time(), '%d %B %Y')`"
documentclass: article
papersize: letter
fontsize: 11pt
output:
  bookdown::pdf_document2
---


```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
## some utility functions, see the source code for details
source("utils_template.R")

## specify the packages needed
pkgs <- "HI"
need.packages(pkgs)

## external data can be read in by regular functions,
## such as read.table or load

## for latex and html output
isHtml <- knitr::is_html_output()
isLatex <- knitr::is_latex_output()
latex <- ifelse(isLatex, '\\LaTeX\\', 'LaTeX')

## specify global chunk options
knitr::opts_chunk$set(fig.width = 5, fig.height = 4, dpi = 300,
                      out.width = "90%", fig.align = "center")

```

# The likelihood function
From the normal mixture example, we know that the likelihood function for the data is
$$
f(x|\theta)=\prod_{i=1}^n [\delta \phi(x_i,\mu_1,\sigma^2_1)+(1-\delta)\phi(x_i,\mu_2,\sigma^2_2)]
$$
where $\phi()$ is the density function of a normal variable. Suppose that prior for $\mu_1$ and $\mu_2$ are $N(0,100)$, that the prior for $1/\sigma_1^2$ and $1/\sigma_2^2$ are $\Gamma(a,b)$ with shape $a=0.5$ and scale $b=10$, and for $\delta$ we use noninformative prior, the unifrom distribution. Further, all the priors are independent. For simplicity, let's write the prior densities for $\mu_1,\mu_2,\sigma_1^2,\sigma_2^2$ are $f_1,f_2,f_3$ and $f_4$, respectively. Thus, the joint posterior distribution for those five parameters is
$$
p(\theta|x) \propto f(x|\theta)f_1(\mu_1)f_2(\mu_2)f_3(\sigma^2_1)f_4(\sigma_2^2)
$$
where $\theta=(\mu_1,\mu_2,\sigma_1^2,\sigma_2^2)$. For applying Gibbs sampling, we have to get the full conditional distribution for each parameter, they are:
\begin{align*}
p(\delta|x,\mu_1,\mu_2,\sigma_1^2,\sigma_2^2) & \propto f(x|\theta) \\
p(\mu_1|x,\delta,\mu_2,\sigma_1^2,\sigma_2^2) & \propto f(x|\theta)f_1(\mu_1) \\
p(\mu_2|x,\delta,\mu_1,\sigma_1^2,\sigma_2^2) & \propto f(x|\theta)f_2(\mu_2) \\
p(\sigma^2_1|x,\delta,\mu_2,\mu_1,\sigma_2^2) & \propto f(x|\theta)f_3(\sigma_1^2) \\
p(\sigma^2_2|x,\delta,\mu_2,\mu_1,\sigma_1^2) & \propto f(x|\theta)f_4(\sigma_2^2). 
\end{align*}

Or we can directly use the relationship:
\begin{align*}
p(\delta|x,\mu_1,\mu_2,\sigma_1^2,\sigma_2^2) & \propto f(x|\theta) \\
p(\mu_1|x,\delta,\mu_2,\sigma_1^2,\sigma_2^2) & \propto f(x|\theta)\\
p(\mu_2|x,\delta,\mu_1,\sigma_1^2,\sigma_2^2) & \propto f(x|\theta)\\
p(\sigma^2_1|x,\delta,\mu_2,\mu_1,\sigma_2^2) & \propto f(x|\theta)\\
p(\sigma^2_2|x,\delta,\mu_2,\mu_1,\sigma_1^2) & \propto f(x|\theta). 
\end{align*}

Then we can conduct Gibbs sampling step by step. For $\delta$, we can just use the rejection sampling method from the example to sample from the conditional density. For all other methods, we can use rejection sampling function \emph{arms} in R-package \emph{HI}. The codes for applying the algorithm is like follows:
```{r}
### Generate data: delta=0.7, mu_1=7, mu_2=10, sigma_1=sigma_2=0.5
delta <- 0.7 
n <- 100
set.seed(254)
u <- rbinom(n, prob = delta, size = 1)
x <- rnorm(n, ifelse(u == 1, 7, 10), 0.5)


mylike <- function(delta,mu1,mu2,sigma1,sigma2,x) {
#  prod(delta * dnorm(x, mu1, sqrt(sigma1)) + (1 - delta) * dnorm(x, mu2, #sqrt(sigma2)))*dnorm(mu1,0,10)*dnorm(mu2,0,10)*dgamma(1/sigma1,0.5,10)*dgamma(1/sigma2,0.5,10)
  prod(delta * dnorm(x, mu1, sqrt(sigma1)) + (1 - delta) * dnorm(x, mu2, sqrt(sigma2)))*dnorm(mu1,0,10)*dnorm(mu2,0,10)*sigma1^(-1.5)*exp(-1/10/sigma1)*sigma2^(-1.5)*exp(-1/10/sigma2)
}

mymcmc5 <- function(niter, init, x){
  res <- matrix(nrow = niter, ncol = 5)
  for (i in 1:niter){
  init.delta <- init[1]
  init.mu1 <- init[2]
  init.mu2 <- init[3]
  init.sigma1 <- init[4]
  init.sigma2 <- init[5]
    
  f0 <- function(x0)  log(mylike(x0,init.mu1,init.mu2,init.sigma1,init.sigma2,x))
  new.delta <- res[i,1] <- arms(init.delta, f0, function(x0) (x0>0)*(x0<1), 1)  

  f1 <- function(x1)  log(mylike(new.delta,x1,init.mu2,init.sigma1,init.sigma2,x))
  new.mu1 <- res[i,2] <- arms(init.mu1, f1, function(x1) (x1>-50)*(x1<100), 1)
  #new.mu1 <- res[i,2] <- arms(rnorm(1,6,1), f1, function(x1) (x1>-50)*(x1<100), 1)
  
  f2 <- function(x2)  log(mylike(new.delta,new.mu1,x2,init.sigma1,init.sigma2,x))
  new.mu2 <- res[i,3] <- arms(init.mu2, f2, function(x2) (x2>-50)*(x2<100),1)
   
  f3 <- function(x3)  log(mylike(new.delta,new.mu1,new.mu2,x3,init.sigma2,x))
  new.sigma1 <- res[i,4] <- arms(init.sigma1, f3, function(x3) (x3>0)*(x3<5), 1)
  
  f4 <- function(x4)  log(mylike(new.delta,new.mu1,new.mu2,new.sigma1,x4,x))
  new.sigma2 <- res[i,5] <- arms(init.sigma2, f4, function(x4) (x4>0)*(x4<5),1)
  
  new.par <- c(new.delta,new.mu1,new.mu2,new.sigma1,new.sigma2)
  init <- new.par
  }
  res
}

niter <- 3000
init <- c(0.5,5.7,9,0.2,0.2)
res <- mymcmc5(niter, init, x)


res <- res[-c(1:1000),]
delta3 <- res[,1]
mu1s <- res[,2]
mu2s <- res[,3]
sigma1s <- res[,4]
sigma2s <- res[,5]

plot(ts(delta3))
plot(ts(mu1s))
plot(ts(mu2s))
plot(ts(sigma1s))
plot(ts(sigma2s))

hist(delta3)
hist(mu1s)
hist(mu2s)
hist(sigma1s)
hist(sigma2s)
```

Or we can directly use the function \emph{arms} to sample five parameters simultaneously, since the underlying method used for multivariate sampling in \emph{arms} is Gibbs sampling.
```{r}
f <- function(xa) log(mylike(xa[1],xa[2],xa[3],xa[4],xa[5],x))
init <- c(0.5,5.7,9,0.2,0.2)
res <- arms(init,f,function(xa) (xa[1]>0)*(xa[1]<1)*(xa[2]>-50)*(xa[2]<100)*(xa[3]>-50)*(xa[3]<100)*(xa[4]>0)*(xa[4]<5)*(xa[5]>0)*(xa[5]<5),3000)

res <- res[-c(1:1000),]
delta3 <- res[,1]
mu1s <- res[,2]
mu2s <- res[,3]
sigma1s <- res[,4]
sigma2s <- res[,5]

plot(ts(delta3))
plot(ts(mu1s))
plot(ts(mu2s))
plot(ts(sigma1s))
plot(ts(sigma2s))

hist(delta3)
hist(mu1s)
hist(mu2s)
hist(sigma1s)
hist(sigma2s)
```












